---
title: "Final Project"
author: "Aarti Ramani"
date: "2023-02-07"
output:
  pdf_document: default
  html_document: default
bibliography: citations.bib
biblio-style: "apalike"
link-citations: true 
---
  
## Project Topic: Airline On-Time Performance, Delays, and Cancellations

## Introduction:

For the final project I would like to analyze airline data to identify different
factors and their effects on a carrier's performance. As a performance measure, 
we would be exploring on-time arrivals, number of cancellations by carrier and 
also explore different reasons for a carrier delay.
Data Science can help identify the major causes of delay and cancellations per carrier. Based on the outcome, carriers can take necessary actions to focus on the problem areas.  

## Problem statement addressed:

This study would benefit airlines by comparing their performances and predicting possibilities of delay based on aircraft/origin/destination and apply corrective measures to reduce cancellations and delays and to improve on-time performance.

## Research Questions 
Following are the topics I would like to focus on as part of this project.

1. Are small carriers reliable in terms of lesser cancellations and delays? 
2. Are the delays seasonal? If yes, which regions are most affected?
3. Does the time of day have any significance on delays?
4. Which carrier has the best on-time performance.
5. Which carrier has the least on-time performance.
6. Identifying the most common cancellation reason for all carriers.
7. Which carrier has the most number of cancellations.
8. Which carrier has the most number of delays.
9. What is the percentage of delays by reason.


## Approach:
I will be performing the following steps:

1. Data analysis - Gathering and understanding different datasets.
2. Data Cleaning and Transforming
3. Merge transformed/cleansed datasets
4. Data visualization/plotting

## Addressing the problem 
Based on the outcomes from data analysis and visualization, 
I would like to identify the following:

- Which carriers are more likely to cause delays or cancellations.
- Which carriers are more reliable in terms of on-time performance.
 

## Datasets

Below data submitted by major carriers to department of transportation (DOT).

 - Flights.csv
 - UniqueCarriers.csv
 - Airports.csv

Data was collected by DOT's Bureau of Transportation Statistics for the year 2022.
The purpose of this data is to analyze airline on-time performance reported by carriers. The datasets has around 40 fields in total of which I will be considering between 15 to 25 columns for analysis. 

## Datasets and Relationships: 
TABLE: **Flights.csv**

Column Name                      | Data Type| Column Description
---------------------------------| ---------| -----------------------------------------
Year	                           |Integer   | Year of extracted flight data
Quarter	                         |Integer   | Quarter 
Month	                           |Integer   | Month of extracted flight data
DayofMonth                       |Integer   | Day of month 
DayOfWeek                        |Integer   | Day of Week
FlightDate                       |Date      | Flight Date
Marketing_Airline_Network        |Character | Marketing Carrier Airline Code
Flight_Number_Marketing_Airline  |Integer   | Marketing Carrier Flight Number
Operating_Airline                |Character | Operating Carrier Airline Code
Tail_Number	                     |Integer   | Operating Carrier Tail Number
Flight_Number_Operating_Airline  |Integer   | Operating Carrier Flight Number
Origin	                         |Character | Origin Airport Code(Airports.csv )
OriginCityName	                 |Character | Origin Airport City Name
OriginState                      |Character | Origin Airport State Code
OriginStateName                  |Character | Origin Airport State Name
OriginWac                        |Integer   | Origin Airport Worlde Area Code
Dest                             |Character | Destination Airport Code(Airports.csv )
DestCityName                     |Character | Destination Airport City Name
DestState                        |Character | Destination Airport State Code
DestStateName                    |Character | Destination Airport State Name
DestWac                          |Integer   | Destination Airport Worlde Area Code
CRSDepTime                       |Integer   | CRS Departure Time (local time: hhmm)
DepTime	                         |Integer   | Actual Departure Time(local time: hhmm)
DepDelay	                       |Integer   | Difference in minutes between scheduled and actual departure time. Early departures show negative numbers.
DepDelayMinutes	                 |Integer   | Difference in minutes between scheduled and actual departure time. Early departures set to 0
DepDel15                         |Integer   | Departure Delay Indicator, 15 Minutes or More (1=Yes)
TaxiOut                          |Integer   | Taxi Out Time, in Minutes
WheelsOff                        |Integer   | Wheels Off Time (local time: hhmm)
WheelsOn                         |Integer   | Wheels On Time (local time: hhmm)
TaxiIn                           |Integer   | Taxi In Time, in Minutes
CRSArrTime                       |Integer   | CRS Arrival Time (local time: hhmm)
ArrTime	                         |Integer   | Actual Arrival Time (local time: hhmm)
ArrDelay	                       |Integer   | Difference in minutes between scheduled and actual arrival time. Early arrivals show negative numbers.
ArrDelayMinutes	                 |Integer   | Difference in minutes between scheduled and actual arrival time. Early arrivals set to 0.   
ArrDel15	                       |Integer   | Arrival Delay Indicator, 15 Minutes or More (1=Yes)
Cancelled	                       |Integer   | Cancelled Flight Indicator (1=Yes)
CancellationCode                 |Integer   | Specifies The Reason For Cancellation
Diverted                         |Integer   | Diverted Flight Indicator (1=Yes)
CRSElapsedTime                   |Integer   | CRS Elapsed Time of Flight, in Minutes
ActualElapsedTime                |Integer   | Elapsed Time of Flight, in Minutes
AirTime                          |Integer   | Flight Time, in Minutes
Flights	                         |Integer   | Number of Flights
Distance                         |Integer   | Distance between airports (miles)
DistanceGroup                    |Integer   | Distance Intervals, every 250 Miles, for Flight Segment
CarrierDelay                     |Integer   | Carrier Delay, in Minutes
WeatherDelay	                   |Integer   | Weather Delay, in Minutes
NASDelay	                       |Integer   | National Air System Delay, in Minutes
SecurityDelay	                   |Integer   | Security Delay, in Minutes
LateAircraftDelay	               |Integer   | Late Aircraft Delay, in Minutes

TABLE: **UniqueCarriers.csv**

Column Name                      | Data Type| Column Description
---------------------------------| ---------| -----------------------------------------
Code                             | Character| Unique Airline Carrier Code
Description                      | Character| Airline Carrier Code Description


TABLE: **Airports.csv** 

Column Name                      | Data Type| Column Description
---------------------------------| ---------| -----------------------------------------
Code                             | Character| Airport Code (IATA)
Description                      | Character| Airport Code Description

\newpage
## Data Considerations:
The following rows will be dropped from the dataset:

- Rows that do not qualify for delay or cancellation
- Rows with missing values for carrier, origin, destination, date and time of    
  departure and arrival will be dropped.
 
## Packages
Following packages are required for the project:

i. dplyr
ii. ggplot2
iii. readr
iv. tidyr

## Data importing and cleaning

### Packages
```{r, message = FALSE, warning = FALSE}

library(readr)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(reshape2) 
```


### How to import and clean my data
### Data importing
```{r, message = FALSE, warning = FALSE}
## Set the working directory to the root of your DSC 520 directory
setwd("C:/Masters/GitHub/Winter2022/Ramani-DSC520/assignments/FinalProject/")


#Merge flight data from Jan through Nov 2022 into a single dataframe
list_of_files <- list.files(path="Data/DOT_Flight_Data/",
                            recursive = TRUE,
                            pattern = "\\.csv$",
                            full.names = TRUE)

merge_flights_df <- readr::read_csv(list_of_files, id = "fl_date") 
nrow(merge_flights_df)
head(merge_flights_df,5)


cancellation_cd <- read_csv(file="Data/DOT/L_CANCELLATION.csv")
nrow(cancellation_cd)
head(cancellation_cd,2)

unique_carrier <- read_csv(file="Data/DOT/L_UNIQUE_CARRIERS.csv") 
nrow(unique_carrier)
head(unique_carrier,2)

airport_cd <- read_csv(file="Data/DOT/L_AIRPORT.csv") 
nrow(airport_cd)
head(airport_cd,2)
```

### Data Transformation and Cleaning
```{r, message = FALSE, warning = FALSE}
# Selecting relevant columns from flights data

carrier_performance_df <- 
  merge_flights_df[c("YEAR","QUARTER","MONTH","DAY_OF_MONTH","DAY_OF_WEEK",
                     "FL_DATE","MKT_UNIQUE_CARRIER","OP_UNIQUE_CARRIER",
                     "ORIGIN","ORIGIN_CITY_NAME","ORIGIN_STATE_ABR","ORIGIN_STATE_NM",
                     "DEST","DEST_CITY_NAME","DEST_STATE_ABR","DEST_STATE_NM",
                     "DEP_DELAY","TAXI_OUT","TAXI_IN",
                     "ARR_DELAY","CANCELLED","CANCELLATION_CODE","DIVERTED",
                     "DISTANCE","CARRIER_DELAY","WEATHER_DELAY","NAS_DELAY",
                     "SECURITY_DELAY","LATE_AIRCRAFT_DELAY")]
 
# Transforming Data

#Cancellation reason in the flight dataset is represented as A, B, C and D.
#Looking up the cancellation code against the cancellation dataset and adding  
#cancellation description to the flight dataframe.

carrier_performance_df$CANCELLATION_REASON <- 
  cancellation_cd$Description[match(carrier_performance_df$CANCELLATION_CODE, 
                                    cancellation_cd$Code)]
 
#Carrier codes in flight dataset are represented as 2 character airline carrier codes.
#Looking up the carrier code against the unique carrier dataset and updating the 
#code by carrier name in the flight dataframe for both operating and marketing carriers. 

carrier_performance_df$MKT_UNIQUE_CARRIER_NAME <- 
  unique_carrier$Description[match(carrier_performance_df$MKT_UNIQUE_CARRIER, 
                                   unique_carrier$Code)]
carrier_performance_df$OP_UNIQUE_CARRIER_NAME <-
  unique_carrier$Description[match(carrier_performance_df$OP_UNIQUE_CARRIER, 
                                   unique_carrier$Code)]

#Airport codes in flight dataset are represented as 3 character airport codes.
#Looking up the airport codes against the airport dataset and updating the 
#airport code by name in the flight dataframe for origin and destination columns.

carrier_performance_df$ORIGIN_AIRPORT <- 
  airport_cd$Description[match(carrier_performance_df$ORIGIN, airport_cd$Code)]
carrier_performance_df$DEST_AIRPORT <- 
  airport_cd$Description[match(carrier_performance_df$DEST, airport_cd$Code)]

#Removing null rows from the dataet
carrier_performance_df <- 
  carrier_performance_df[,colSums(is.na(carrier_performance_df))<nrow(carrier_performance_df)]
 
#Updating blank arrival delay to 0
carrier_performance_df[is.na(carrier_performance_df$ARR_DELAY),]$ARR_DELAY = 0

# Add a new column with the performance status
carrier_performance_df <- carrier_performance_df %>% mutate(
  Performance = case_when(
    CANCELLED==1~"Cancelled",
    DIVERTED==1~"Diverted",
    ARR_DELAY<=15~"On-Time",
    ARR_DELAY>15~"Delayed"))

```
### What does the final data set look like?
```{r}
nrow(carrier_performance_df)
names(carrier_performance_df)

```
\newpage
### What information is not self-evident?
I would like to see if there are weather delays or cancellations specific to 
a time of year. If yes, I would like to see if it can be isolated to a particular 
airport or carrier.Also, I am hoping to evaluate the reason reported. 
Was it reported as a weather delay or a NAS delay.
This would probably give an option to see which carrier has reported the most number of NAS delays during bad weather.

### What are different ways you could look at this data ?
I would like to perform the following:

1. Percentages of flights in and out per airline.

2. Percentages of flights in vs delayed per airline.

3. Identify the correlations between variables and perform further analysis based on the outcomes.

### Do you plan to slice and dice the data? 

Currently I am splitting dataset into 2 categories. 

1. no cancellations and delays (on-time performace)

2. cancellations or delays 

```{r,message = FALSE, warning = FALSE}

carrier_on_time_performance_df <-
  carrier_performance_df[(is.na(carrier_performance_df$CANCELLATION_CODE)&
                            carrier_performance_df$ARR_DELAY <= 15),]

carrier_cancel_or_delay_df <-
  carrier_performance_df[!(is.na(carrier_performance_df$CANCELLATION_CODE)&
                            carrier_performance_df$ARR_DELAY <= 15),]
nrow(carrier_on_time_performance_df) + nrow(carrier_cancel_or_delay_df)

```
Splitting cancelled and delayed data.

Delayed Dataset
```{r,message = FALSE, warning = FALSE}
carrier_delay_df <- 
  carrier_cancel_or_delay_df[carrier_cancel_or_delay_df$ARR_DELAY > 15,]
nrow(carrier_delay_df)
```
Cancelled Dataset
```{r,message = FALSE, warning = FALSE}
carrier_cancelled_df <- 
  carrier_cancel_or_delay_df[!is.na(carrier_cancel_or_delay_df$CANCELLATION_CODE),]
nrow(carrier_cancelled_df)  
```

```{r,message = FALSE, warning = FALSE}
nrow(carrier_delay_df)
nrow(carrier_cancelled_df)
nrow(carrier_on_time_performance_df)
nrow(carrier_performance_df)
```
Apart from the above, since the number of rows are very high at this point, 
I will  narrow my research to flights origination from major 20 airports. 

```{r,message = FALSE, warning = FALSE}
carrier_performance_df <- 
  carrier_performance_df[carrier_performance_df$ORIGIN == "ORD" 
                                 | carrier_performance_df$ORIGIN == "ATL" 
                                 | carrier_performance_df$ORIGIN == "DFW" 
                                 | carrier_performance_df$ORIGIN == "DEN" 
                                 | carrier_performance_df$ORIGIN == "EWR" 
                                 | carrier_performance_df$ORIGIN == "LAX" 
                                 | carrier_performance_df$ORIGIN == "IAH" 
                                 | carrier_performance_df$ORIGIN == "PHX" 
                                 | carrier_performance_df$ORIGIN == "DTW" 
                                 | carrier_performance_df$ORIGIN == "SFO" 
                                 | carrier_performance_df$ORIGIN == "LAS" 
                                 | carrier_performance_df$ORIGIN == "DEN" 
                                 | carrier_performance_df$ORIGIN == "ORD" 
                                 | carrier_performance_df$ORIGIN == "JFK" 
                                 | carrier_performance_df$ORIGIN == "CLT" 
                                 | carrier_performance_df$ORIGIN == "LGA" 
                                 | carrier_performance_df$ORIGIN == "MCO" 
                                 | carrier_performance_df$ORIGIN == "MSP" 
                                 | carrier_performance_df$ORIGIN == "BOS" 
                                 | carrier_performance_df$ORIGIN == "PHL",]

nrow(carrier_performance_df)


carrier_delay_df <- 
  carrier_delay_df[carrier_delay_df$ORIGIN == "ORD" 
                   | carrier_delay_df$ORIGIN == "ATL"
                   | carrier_delay_df$ORIGIN == "DFW" 
                   | carrier_delay_df$ORIGIN == "DEN" 
                   | carrier_delay_df$ORIGIN == "EWR" 
                   | carrier_delay_df$ORIGIN == "LAX" 
                   | carrier_delay_df$ORIGIN == "IAH" 
                   | carrier_delay_df$ORIGIN == "PHX" 
                   | carrier_delay_df$ORIGIN == "DTW" 
                   | carrier_delay_df$ORIGIN == "SFO" 
                   | carrier_delay_df$ORIGIN == "LAS" 
                   | carrier_delay_df$ORIGIN == "DEN" 
                   | carrier_delay_df$ORIGIN == "ORD" 
                   | carrier_delay_df$ORIGIN == "JFK" 
                   | carrier_delay_df$ORIGIN == "CLT" 
                   | carrier_delay_df$ORIGIN == "LGA" 
                   | carrier_delay_df$ORIGIN == "MCO" 
                   | carrier_delay_df$ORIGIN == "MSP" 
                   | carrier_delay_df$ORIGIN == "BOS" 
                   | carrier_delay_df$ORIGIN == "PHL",]
nrow(carrier_delay_df)


carrier_cancelled_df <- 
  carrier_cancelled_df[carrier_cancelled_df$ORIGIN == "ORD" 
                       | carrier_cancelled_df$ORIGIN == "ATL" 
                       | carrier_cancelled_df$ORIGIN == "DFW" 
                       | carrier_cancelled_df$ORIGIN == "DEN" 
                       | carrier_cancelled_df$ORIGIN == "EWR" 
                       | carrier_cancelled_df$ORIGIN == "LAX" 
                       | carrier_cancelled_df$ORIGIN == "IAH" 
                       | carrier_cancelled_df$ORIGIN == "PHX" 
                       | carrier_cancelled_df$ORIGIN == "DTW" 
                       | carrier_cancelled_df$ORIGIN == "SFO" 
                       | carrier_cancelled_df$ORIGIN == "LAS" 
                       | carrier_cancelled_df$ORIGIN == "DEN" 
                       | carrier_cancelled_df$ORIGIN == "ORD" 
                       | carrier_cancelled_df$ORIGIN == "JFK" 
                       | carrier_cancelled_df$ORIGIN == "CLT" 
                       | carrier_cancelled_df$ORIGIN == "LGA" 
                       | carrier_cancelled_df$ORIGIN == "MCO" 
                       | carrier_cancelled_df$ORIGIN == "MSP" 
                       | carrier_cancelled_df$ORIGIN == "BOS" 
                       | carrier_cancelled_df$ORIGIN == "PHL",]

nrow(carrier_cancelled_df)
 
carrier_on_time_performance_df <- 
  carrier_on_time_performance_df[carrier_on_time_performance_df$ORIGIN == "ORD" 
                                 | carrier_on_time_performance_df$ORIGIN == "ATL" 
                                 | carrier_on_time_performance_df$ORIGIN == "DFW" 
                                 | carrier_on_time_performance_df$ORIGIN == "DEN" 
                                 | carrier_on_time_performance_df$ORIGIN == "EWR" 
                                 | carrier_on_time_performance_df$ORIGIN == "LAX" 
                                 | carrier_on_time_performance_df$ORIGIN == "IAH" 
                                 | carrier_on_time_performance_df$ORIGIN == "PHX" 
                                 | carrier_on_time_performance_df$ORIGIN == "DTW" 
                                 | carrier_on_time_performance_df$ORIGIN == "SFO" 
                                 | carrier_on_time_performance_df$ORIGIN == "LAS" 
                                 | carrier_on_time_performance_df$ORIGIN == "DEN" 
                                 | carrier_on_time_performance_df$ORIGIN == "ORD" 
                                 | carrier_on_time_performance_df$ORIGIN == "JFK" 
                                 | carrier_on_time_performance_df$ORIGIN == "CLT" 
                                 | carrier_on_time_performance_df$ORIGIN == "LGA" 
                                 | carrier_on_time_performance_df$ORIGIN == "MCO" 
                                 | carrier_on_time_performance_df$ORIGIN == "MSP" 
                                 | carrier_on_time_performance_df$ORIGIN == "BOS" 
                                 | carrier_on_time_performance_df$ORIGIN == "PHL",]

nrow(carrier_on_time_performance_df)
```

### How could you summarize your data to answer key questions? 
Calculating the correlation and covariance are great ways to summarize my data to answer key questions. Results from the summary function would also help. 
In addition, finding the maximum, minimum, mean, and median values for delays will provide some more information.


## Plots & Tables
Plots that I would like to explore:

i. Scatter plot 
ii. Pie chart   
iii. Histogram
iv. Boxplot
    
I will create tables with the following data:

A summary table of on-time performance, delays, and cancellations per carrier. 

\newpage

### What types of plots and tables will help you to illustrate the findings to your questions?

### BOXPLOT

```{r,message = FALSE, warning = FALSE}
#Boxplot for checking Outliers in Airline Delays
ggplot(carrier_delay_df, aes(x=ARR_DELAY, y=OP_UNIQUE_CARRIER))+ 
  geom_boxplot(outlier.colour="orange", outlier.shape=16) + 
  labs(title ="Airline Delay by Operating Carrier",
       y = "Carrier Airline", 
       x = "Arrival Delay (minutes)") 

```

We can see many outliers in the delay dataset which can be omitted.
We can skip rows with arrival delay greater than 2000 minutes. Since 90% of the data falls below 2000 minutes. Also, we can see more outliers for the carriers American Airlines, Mesa,Skywest and Republic airlines.

\newpage

```{r,message = FALSE, warning = FALSE}
# Box Plot for Airline Delays by Delay Reasons

carrier_delay_df_mod <- melt(carrier_delay_df, id.vars='ARR_DELAY',
                  measure.vars=c("CARRIER_DELAY","WEATHER_DELAY","NAS_DELAY",
                                 "SECURITY_DELAY","LATE_AIRCRAFT_DELAY"))
# creating a plot
ggplot(carrier_delay_df_mod) + 
  geom_boxplot(aes(x=ARR_DELAY, y=value, color=variable)) +
  xlab("Arrival Delay in minutes") + ylab("Value") + ggtitle("Airline Delays by Delay Reasons") 

```
It appears there are more delays due to carrier and late aircrafts than weather, NAS or security delays.
We can also see a few outliers in the delay data that can be removed for the analysis.

\newpage

### HISTOGRAM

```{r,message = FALSE, warning = FALSE}
# Histogram to  the Cancellations due to weather
ggplot(carrier_cancelled_df, aes(x=MONTH, fill=CANCELLATION_REASON)) + 
  geom_histogram(bins=25) + xlab("Month") + ylab("Count") + 
  scale_x_continuous(breaks = seq(1, 12, by = 1)) + 
  ggtitle("Airline Cancellations per month")

```

We can see that weather is the major reason for cancellation, especially in the winter (January and February). Carrier cancellations are also high during this period.

\newpage

```{r,message = FALSE, warning = FALSE}
# Histogram for Airline Cancellations
carrier_performance_df$CANCELLED_F <-factor(carrier_performance_df$CANCELLED)
 
ggplot(data=carrier_performance_df,
       aes(x=OP_UNIQUE_CARRIER,fill=CANCELLED_F))+
  geom_histogram(stat = "Count",position = "dodge", col="red", bins = 15) +
  labs(title="Cancellations by Operating Airlines") + 
  labs(x="Airline", y="Number of Flights") +scale_x_discrete() +
  scale_y_log10() +
  labs(fill='Cancelled or Not')

```

At this time, it appears American, Delta, United, Republic Airways etc., have more cancellations.
However, we cannot conclude this to be true, since they also have more fleets in comparison to other low cost airlines. I would like to redo this plot with percentage of cancellations per airline.

\newpage

### PIE CHART

```{r,message = FALSE, warning = FALSE}
#Pie Chart for Over all performance
flight_stats <- carrier_performance_df %>% count(Performance)

#Average number of delayed flights per month since 2003
avg_flights_cancelled <- flight_stats[flight_stats$Performance == "Cancelled",]$n
avg_flights_diverted <- flight_stats[flight_stats$Performance == "Diverted",]$n
avg_flights_on_time <- flight_stats[flight_stats$Performance == "On-Time",]$n
avg_flights_delayed <- flight_stats[flight_stats$Performance == "Delayed",]$n
 

flights_values = 
  c(avg_flights_cancelled,avg_flights_delayed,avg_flights_diverted,avg_flights_on_time)
flights_labels = c("Cancelled","Delayed","Diverted","On Time")

#Calculating the percent each will be of the total
piepercent<- round(100*flights_values/sum(flights_values), 1) 
  
pie(flights_values, labels = piepercent, main = "Airlines Performance Percentages",
    col = brewer.pal(9, "Pastel1"))

legend("topright",flights_labels, cex = 0.8,fill = brewer.pal(9, "Pastel1"))
```
\newpage

### What do you not know how to do right now that you need to learn to answer your questions?
I would like to learn more on the machine learning concepts to use in my final project.


### Do you plan on incorporating any machine learning techniques to answer your research questions? Explain.
Not at this time, but would like to consider incorporating them based on week 11 and 12 learnings.

## Questions
It is unclear if I would be able to recommend the right area of focus for better performance, to the airlines. 

For example: 
If the majority of delays are due to NAS - National Air System Delay, 
it could mean there was an issue in one or more areas such as mechanical, crew, airport operations etc. 
I would need to identify another dataset that logs the maintenance or operational issues by carrier. This information could be hard to get as it is carrier specific and probably not allowed to be made public.


## Citations
[@airlinedataset]
